Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Introduction.tex

Formal verification of cache coherence protocols has traditionally been done
using model checking. That family of analyses significantly restricts the size as
well as complexity of the system that has to be verified; most model-checking
based approaches can not handle more than one level of hierarchy. To the best
of our knowledge, the largest system that has been verified using
model checking has only 12 processors, in an era when many observers are
predicting mainstream systems with hundreds of processor cores.
\xxx{citation:
http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
data from their related work must cite)}.

@@

The main contribution of this paper is \textbf{the first verification of a
cache coherence protocol that is parametric over an infinite family of
memory hierarchies}.  Our proof is implemented in the Coq proof assistant and
checked via a very small trusted code base.  We also want to draw attention to
a few key elements of our methodology that should be of interest for future
verifications of this kind, both through literal code reuse and through insights
about key invariants that should generalize across protocols.
\begin{enumerate}
\item We give a \emph{precise formal definition of a cache-coherent system} as one
that obeys \emph{store atomicity}. The definition is abstract in that it does not
rely on any hardware implementation.
\item We provide a \emph{Coq-based proof of the directory-based MSI cache coherence
protocol} with inclusive caches and no silent drops.
\item We present \emph{a set of invariants} that are not restricted to the MSI
protocol but which imply store atomicity. Thus, other protocols will also
satisfy store atomicity as long as they obey the invariants.
\item We present what we believe to be \emph{the key insight behind the
correctness of the MSI protocol}. We believe that this property is true also of
other protocols like MESI, MOSI, and MOESI, enabling reuse of portions of our
MSI proof for their verification.
\end{enumerate}
%In practice, verification proceeds in
%two steps. First, a cache coherence protocol is constructed and verified. This
%protocol generally directly represents many of the key concurrent aspects of an
%implementation (\eg{} reads and writes to the system are broken into local
%atomic micro-operations) but ignores implementation details (\eg{} buffer
%sizing and memory access times). Second, once the protocol has been verified,
%it is handed to hardware experts who fill in the missing implementation details
%and realize it at the Register Transfer Level (RTL) of gates, wires, and
%registers, via further breaking down of local micro-operations to make them
%realizable in hardware. This effort is not a one-time translation, as
%performance properties are not easily estimated; a significant amount of
%performance testing is requires to select a good implementation. 
%
%\adamc{I don't understand what this paragraph is saying.}
%From a methodological perspective, this two-step process leaves much to be
%desired. The significant effort of verification is effectively lost by the
%concurrent non-trivial tasks of translation from the protocol representation to
%a hardware description and the necessary refinement to add missing
%implementation details to the protocol.

Protocols, post verification, are often described via Term Rewrite Systems
(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
systems at an interesting point on the spectrum between declarative and
operational methods.  It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
hardware description language based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly implemented into
efficient hardware. In this way one reduces the correctness of the
implementation to the correctness of the protocol, with no need for separate
reasoning about an implementation. While the current Bluespec compiler does not
support some implementation elements that are common in practice (\eg{}
fairness guarantees, atomic actions running over multiple clock cycles), recent
work has provided key insights toward lifting the
restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
evolutions represent a very promising platform for productive implementation of
hardware components, including cache-coherent shared memories, starting from a
description designed to support verification. Our model of the MSI protocol
in Coq is described as an inductively defined set of transitions that is
directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
is a direct transliteration of the MSI protocol specified in Bluespec,
ensuring that the final hardware synthesized from the specification is formally
verified.

@@

\noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
formal definition of a cache-coherent system as one that obeys store atomicity.
We describe memory subsystems as mathematical functions and define store
atomicity as a property of the input/output pairs of the memory subsystem
function. We describe our MSI protocol and its representation as a
transition relation in Coq in Section~\ref{sec:System}. In
Section~\ref{safety} we discuss the structuring of our proof. We
present the invariants that are sufficient to prove store atomicity and give
the key insight that enables these invariants to be proved.
%We discuss some refinements in
%Section~\ref{Sec:Refinements} and conclude with a general discussion
%of the applicability of our technique.

>>>>

Formal verification of cache coherence protocols has traditionally been done
using model-checking based approaches. This significantly restricts the size as
well as complexity of the system that has to be verified; most model-checking
based approaches cannot handle more than one level of hierarchy. To the best of
our knowledge, the largest system that has been verified using model-checking
based approaches has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era
when many observers are predicting mainstream systems with hundreds of processor
cores.

@@

The main contribution of this paper is \textbf{the first verification of a
cache coherence protocol that is parametric over an infinite family of
memory hierarchies}.  Our proof is implemented in the Coq proof assistant and
checked via a very small trusted code base.  We also want to draw attention to
a few key elements of our methodology that should be of interest for future
verifications of this kind, both through literal code reuse and through insights
about key invariants that should generalize across protocols.
\begin{enumerate}

\item We give a \emph{precise formal definition of a cache-coherent system} as one
that obeys \emph{store atomicity}. The definition is abstract in that it does not
rely on any hardware implementation.
\item We provide a \emph{Coq-based proof of the directory-based MSI cache coherence
protocol} with inclusive caches and no silent drops.
\item We present \emph{a set of invariants} that are not restricted to the MSI
protocol but which imply store atomicity. Thus, other protocols will also
satisfy store atomicity as long as they obey the invariants.
\item We present what we believe to be \emph{the key insight behind the
correctness of the MSI protocol}. We believe that this property is true also of
other protocols like MESI, MOSI, and MOESI, enabling reuse of portions of our
MSI proof for their verification.

\end{enumerate}

%In practice, verification proceeds in
%two steps. First, a cache coherence protocol is constructed and verified. This
%protocol generally directly represents many of the key concurrent aspects of an
%implementation (\eg{} reads and writes to the system are broken into local
%atomic micro-operations) but ignores implementation details (\eg{} buffer
%sizing and memory access times). Second, once the protocol has been verified,
%it is handed to hardware experts who fill in the missing implementation details
%and realize it at the Register Transfer Level (RTL) of gates, wires, and
%registers, via further breaking down of local micro-operations to make them
%realizable in hardware. This effort is not a one-time translation, as
%performance properties are not easily estimated; a significant amount of
%performance testing is requires to select a good implementation. 
%
%\adamc{I don't understand what this paragraph is saying.}
%From a methodological perspective, this two-step process leaves much to be
%desired. The significant effort of verification is effectively lost by the
%concurrent non-trivial tasks of translation from the protocol representation to
%a hardware description and the necessary refinement to add missing
%implementation details to the protocol.

Protocols, post verification, are often described via Term Rewrite Systems (TRS)
\adamc{citation for TRSes?}, which apply equational logic to describe systems at
an interesting point on the spectrum between declarative and operational
methods.  It was shown by Dave \etal{}~\cite{DNA:CoherenceImplementation} that
such a protocol, after its verification, can be transliterated into
Bluespec~\cite{Bluespec:TFRG}, a hardware description language based on Guarded
Atomic Actions, a restriction of TRS~\cite{Hoe:TCAD}. Bluespec specification can
be directly implemented into efficient hardware. In this way one reduces the
correctness of the implementation to the correctness of the protocol, with no
need for separate reasoning about an implementation. While the current Bluespec
compiler does not support some implementation elements that are common in
practice (\eg{} fairness guarantees, atomic actions running over multiple clock
cycles), recent work has provided key insights toward lifting these
restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
evolutions represent a promising platform for productive implementation of
hardware, including cache-coherent shared memories, starting from a
verification-empowering description.

Our model of the MSI protocol in Coq is described as an inductively defined set
of transitions which is directly translatable into Guarded Atomic Actions. In
fact, our Coq-based model is a direct transliteration of the MSI protocol
specified in Bluespec.  This ensures that the final hardware synthesized from
the specification is formally verified.

@@

\noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
formal definition of a cache-coherent system as one that obeys store atomicity.
We describe memory subsystems as mathematical functions and define store
atomicity as a property of the input/output pairs of the memory subsystem
function. We describe our MSI protocol and its representation as a
transition relation in Coq in Section~\ref{sec:System}. In
Section~\ref{safety} we discuss the structuring of our proof. We
present the invariants that are sufficient to prove store atomicity and give
the key insight that enables these invariants to be proved.

%We discuss some refinements in
%Section~\ref{Sec:Refinements} and conclude with a general discussion
%of the applicability of our technique.

\adamc{This might be a good place to include a link to the Coq source code.}

### 2014-02-06 paper/Intuition.tex
spell in later half of paper

\section{Proof that the MSI Protocol Preserves Store Atomicity}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to the MSI protocol;
any other protocol that obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight that we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the examples of MESI,
MOSI and MOESI in Section \ref{mosi}.
\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

@@

%ndave: This needs some repointing
We present what we believe to be the key insight for proving any cache-coherence protocol.
In fact, the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address.
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address.
\item The state or directory entry changes for an address only on sending and
receiving a response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

\vspace{-.1in}

@@

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.

>>>>

\section{Proof that the MSI Protocol Preserves Store Atomicity}
\label{Sec:Safety}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to MSI protocol
alone; any other protocol which obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight which we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the example of MESI,
MOSI and MOESI in Section~\ref{sec:Refinements}.
\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

@@

We present what we believe to be the key insight for proving any cache-coherence protocol.
In fact, the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address if and only if it sends a response for that address
\item A parent upgrades the directory entry for an address for one of
  its children if and only if it sends a response for that child and
  address
\item The state or directory entry changes for an address only on sending and
receiving a response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}
\vspace{-.1in}

@@

Figure~\ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure~\ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section~\ref{secondLevel} requires using more
properties of the MSI transistions of Figure~\ref{trans}, though it is
straightforward. Analysing a parent-child pair will suffice in proving
Invariants~\fullref{localCompatibility1} and
\fullref{localCompatibility2}. Proof of Invariant~\fullref{noTransitWrite} is
similar to the proof of Invariant~\fullref{nonAncestorCompatible}.

### 2014-02-06 paper/System.tex

\section{MSI-based cache coherence protocol}
\label{sec:System}

We have chosen as our concrete example a hierarchical MSI
protocol~\cite{MSI}. Logically, the system is organized in a tree
structure, with the root connected to the main memory and each leaf interfaced
with a processor (Figure~\ref{hier}). The system is parameterized by the shape
of the tree, and each node can have any number of children independent of the
number in other nodes. In the following, $\leaf(c)$ indicates
that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
$c$.

@@

Input and output happens at the leaves in the form of memory requests
issued from and responses issued to the associated processor. These will be
represented by two mappings, one from $\mathbb{N}$ to requests and the other
from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
sequence number of the next request to process; it is set to $0$ initially.
\xxx{Fixed???? \adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
\adamc{I hope the diagram explains what these mappings mean, and in general what
the tree structure really means.}}

Each node maintains a permission state for each address $a$, $c.\state[a]$, in
addition to the data associated with the address, $c.\data[a]$. The permission
state denotes which operations may be performed on the data. An address may be
in the \emph{read-write} state \Mo{} (for ``modifiable''),
the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
\In{} (indicating that it can be neither read nor written). The options form a natural
total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
of these permission states is straightforward: a load request can be processed
only if the leaf is at least in the \Sh{} state, and a store request can be
processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
an \emph{upper bound} on the permission level of any cache in the associated subtree.
If a cache does not include a particular location,
we consider that address to have permission \In.  Initially, all addresses in
each cache are in the \In{} state, save for those in the root cache (\ie{} main memory),
which are in the \Mo{} state.

@@

We have dedicated channels between each parent-child pair. The three channels
are indexed by the name of the child that uses that channel (denoted by
$\textit{ch}[c]$ for the \textit{ch} channel associated
with cache $c$).  Figure~\ref{format} shows the message formats
transmitted in the 3
channels between caches. In each of the channels $ch$, a message $m$
can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
channels can be examined for the first element ($\first(ch)$) or for
availability of any message in the channel
($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch,
\neg \avail(ch)$, initially.

\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{\Reqcp}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{\Respcp}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{\Mpc}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches}
\label{format}
\end{figure}

@@

We first give an intuition on how the transitions work in the common case.  A
cache may spontaneously choose to upgrade its state, in which case it sends an
upgrade request to its parent. The parent then makes a local decision on whether
to send a response to the requesting child or not (based on its directory
approximation, and its own state). If its own state is lower the requested
upgrade, then it can not handle the request. The parent must instead, decide to
upgrade its state -- the transition for doing so has nothing to do with the
incoming request.  Once the parent's state is not lower than the requested
upgrade, it makes sure that the rest of its children are ``compatible'' with the
requested upgrade. If not, the parent must, send requests to the incompatible
children to downgrade -- again the transition for doing so has nothing to do
with the incoming request. If the children are compatible and its own state is
higher, then the request can be handled and a response is sent. Each of the
transitions is explained in greater detail below.

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c,
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child receives the response from the parent (ChildRecvResp)
and sets its state, resetting its wait state if the response contains the
desired upgrade.

@@

Because of the presence of voluntary response, a child could have already
downgraded to a state required by a downgrade request from its parent. In such
cases, the child simply drops the downgrade request (ChildDropReq).

@@

Note that the atomic transitions in Figure~\ref{trans} access only local states,
\ie{} a transition can read or write states only corresponding to a single cache
and/or the channel the cache is connected to. This restriction allows such a
system to be implemented directly into hardware. Bluespec SystemVerilog (BSV),
for instance, directly converts these transitions into efficient synchronous
hardware. Though these transitions logically happen one by one, BSV executes
several transitions simultaneously, as long as they do not access the same
state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
Sometimes, these transitions (both the guards and the actions) may not be
amenable to single cycle implementations in hardware. For example, writing a
register array can actually take several hardware clock cycles, though
logically it is a single transition. Karczmarek \etal \cite{Karczmarek} has
provided a scheme to convert atomic transitions into synchronous hardware in
which each transition can span several clock cycles.

>>>>

\section{MSI-Based Cache Coherence Protocol}
\label{sec:System}

We have chosen as our concrete example a hierarchical MSI
protocol~\cite{MSI}. Logically, the system is organized in a tree
structure, with the root connected to the main memory and each leaf interfaced
with a processor (Figure~\ref{hier}). The system is parameterized by the shape
of the tree, and each node can have any number of children.
In the following, $\leaf(c)$ indicates
that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
$c$.

@@

Input and output happen at the leaves in the form of memory requests
issued from and responses issued to the associated processors. These will be
represented by two mappings, one from $\mathbb{N}$ to requests and the other
from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
sequence number of the next request to process; it is set to $0$ initially.
\adamc{Add processors and main memory, labeled explicitly, to the diagram?}

Each node maintains a permission state for each address $a$, $c.\state[a]$, in
addition to the data associated with the address, $c.\data[a]$. The permission
state denotes which operations may be performed on the data. An address may be
in the \emph{read-write} state \Mo{} (for ``modifiable''),
the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
\In{} (indicating that it may be neither read nor written). The options form a natural
total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
of these permission states is straightforward: a load request may be processed
only if the leaf is at least in the \Sh{} state, and a store request may be
processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
an \emph{upper bound} on the permission level of any cache in the associated subtree.
If a cache does not include a particular location,
we consider that address to have permission \In.  Initially, all addresses in
each cache are in the \In{} state, save for those in the root cache (\ie{} main memory),
which are in the \Mo{} state.

@@

We have dedicated channels between each parent-child pair. The three channels
are indexed by the name of the child that uses that channel (denoted by
$\textit{ch}[c]$ for the \textit{ch} channel associated
with cache $c$).  Figure~\ref{format} shows the message formats
transmitted in the 3
channels between caches. In each of the channels $ch$, a message $m$
can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
channels can be examined for the first element ($\first(ch)$) or for
availability of any message in the channel
($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch. \;
\neg \avail(ch)$, initially.

\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{\Reqcp}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{\Respcp}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{\Mpc}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches
\adamc{Names here don't match those in the preceding text.}}
\label{format}
\end{figure}

@@

We first give an intuition on how the transitions work in the common
case.  A cache can decide, unprovoked, to upgrade its state, in which
case it sends an upgrade request to its parent. The parent then makes
a local decision on whether to send a response to the requesting child
or not (based on its directory approximation, and its own state). If
its own state is lower the requested upgrade, then it cannot handle
the request. The parent must instead decide to upgrade its state --
the transition for doing so has nothing to do with the incoming
request.  Once the parent's state is not lower than the requested
upgrade, it makes sure that the rest of its children are
``compatible'' with the requested upgrade. If not, the parent must
send requests to the incompatible children to downgrade -- again the
transition for doing so has nothing to do with the incoming
request. If the children are compatible and its own state is higher,
then the request can be handled and a response is sent. Each of the
transitions is explained in greater detail below.

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c. \;
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child receives the response from the parent (ChildRecvResp)
and sets its state, resetting its wait state if the response contains the
desired upgrade.

@@

Because of the presence of voluntary responses, a child could have already
downgraded to a state required by a downgrade request from its parent. In such
cases, the child simply drops the downgrade request (ChildDropReq).

@@

Note that the atomic transitions in Figure~\ref{trans} access only local states,
\ie{} a transition can read or write states only corresponding to a single cache
and/or the channel the cache is connected to. This restriction allows such a
system to be implemented directly into hardware. Bluespec SystemVerilog (BSV),
for instance, directly converts these transitions into efficient synchronous
hardware. Though these transitions logically happen one by one, BSV executes
several transitions simultaneously, as long as they do not access the same
state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
Sometimes, these transitions (both the guards and the actions) may not be
amenable to single-cycle implementations in hardware. For example, writing a
register array can actually take several hardware clock cycles, though
logically it is a single transition. Karczmarek \etal \cite{Karczmarek} has
provided a scheme to convert atomic transitions into synchronous hardware in
which each transition can span several clock cycles.
